---
output: github_document 
---

# bisonpictools

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/poissonconsulting/bisonpictools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/poissonconsulting/bisonpictools/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/poissonconsulting/bisonpictools/graph/badge.svg?token=znqSXiltZo)](https://codecov.io/gh/poissonconsulting/bisonpictools)
<!-- badges: end -->

The goal of `bisonpictools` is to facilitate the visualization and analysis of camera trap data for Wood Bison herds.
The package includes functions to check the correct formatting of data, visualize the data, manipulate and analyse the data using a complex custom-built Bayesian model, and generate predictions of abundances, survival and fecundity rates, and population ratios.

## bisonpic Suite

`bisonpictools` is one of three packages in the **bisonpic** suite of tools:

- [`bisonpictools`](https://github.com/poissonconsulting/bisonpictools)
- [`shinybisonpic`](https://github.com/poissonconsulting/shinybisonpic)
- [`runbisonpic`](https://github.com/poissonconsulting/runbisonpic)

`bisonpictools` provides the underlying visualization and analysis functionality; `shinybisonpic` is a web-based Shiny app that provides a Graphical User Interface to facilitate data visualization and exploration, and `runbisonpic` is a local-based Shiny app to set up and run an analysis.

## Installation

To install the latest development version of `bisonpictools`:

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("poissonconsulting/bisonpictools")
```

And to load the package into the current R session:

```{r}
library(bisonpictools)
```

## Data

`bisonpictools` accepts data in a specific format for wood bison camera trap locations and events, as well as census and proportion of calf estimates from aerial surveys of the herd.
There is a downloadable excel template in each of the Shiny apps that aids the user to populate their data appropriately.

```{r}
# Example data for locations of camera traps
bpt_location_data
```

```{r}
# Example data for camera trap events
bpt_event_data
```

```{r}
# Example data for census estimate from aerial surveys
bpt_census_data
```

```{r}
# Example data for proportion of calves estimate from aerial surveys
bpt_proportion_calf_data
```

## Data Visualization

Use the `bpt_plot_ratios()` function to visualize ratios in the classified counts from the camera trap events over time.
The ratio plotted is `numerator:(denominator + numerator)` to avoid infinite values.

For example, the adult cow:(bull + cow) ratio plotted over all years and locations:
``` {r, fig.dim = c(12, 8)}
library(bisonpictools)
bpt_plot_ratios(
  bpt_event_data,
  bpt_location_data,
  numerator = "fa",
  denominator = "ma"
)
```

## Data Analysis

The `bpt_analyse()` function runs the analysis.
The following example uses the built-in example data in the demonstrative "quick" analysis mode.

```{r, eval = FALSE}
analysis <- bpt_analyse(
  event_data = bpt_event_data,
  location_data = bpt_location_data,
  census_data = bpt_census_data,
  proportion_calf_data = bpt_proportion_calf_data,
  nthin = 1L,
  analysis_mode = "quick"
)
```
```{r, echo = FALSE}
analysis <- readRDS(file = system.file(package = "bisonpictools", "test-objects/analysis.RDS"))
```
## View model estimates

Use the `bpt_coefficient_table()` function to view the model estimates.

```{r}
bpt_coefficient_table(analysis)
```

## Make predictions

Predictions of the total and the class-wise abundances, the fecundity and survival rates, and several ratios of interest are derived from the posterior distribution of the estimated parameters.
These can be visualized in tables using the functions with the `bpt_predict` prefix:

```{r}
# E.g., Predict total abundance
bpt_predict_abundance_total(analysis)
```

```{r, eval = FALSE}
# Can predict other values using the following functions
bpt_predict_abundance_class(analysis) # Predicts abundance by class
bpt_predict_survival(analysis) # Predicts survival rates
bpt_predict_fecundity(analysis) # Predicts fecundity rate and proportion of reproductive cows
bpt_predict_ratios(analysis) # Predicts population ratios
```

Each of the above predictions can be plotted in a figure using the `bpt_plot_predictions()` function.

```{r}
# E.g., Plot predicted abundances by class
bpt_plot_predictions(analysis, prediction = "abundance-class")
```

The other predictions can be plotted by executing the following code:
```{r, eval = FALSE}
# Plot total abundance
bpt_plot_predictions(analysis, prediction = "abundance-total")
# Plot survival rates
bpt_plot_predictions(analysis, prediction = "survival")
# Plot fecundity rates
bpt_plot_predictions(analysis, prediction = "fecundity")
# Plot ratios
bpt_plot_predictions(analysis, prediction = "ratios")
```

## Information

Additional information is available from the [`bisonpictools` website](https://poissonconsulting.github.io/bisonpictools/).
The [user guide](https://poissonconsulting.github.io/bisonpictools/user-guide.html), and the ["Getting Started"](https://poissonconsulting.github.io/bisonpictools/bisonpictools.html) and ["Analytical Methods"](https://poissonconsulting.github.io/bisonpictools/bisonpicmethods.html) vignettes may be particularly useful.

## Contribution

Please report any [issues](https://github.com/poissonconsulting/bisonpictools/issues).

## Code of Conduct

Please note that the `bisonpictools` project is released with a [Contributor Code of Conduct](https://www.contributor-covenant.org/version/2/1/code_of_conduct/).
By contributing to this project, you agree to abide by its terms.

```{r, results = "asis", echo = FALSE}
cat(bpt_licensing_md())
```
