---
title: Analytic Methods for Estimating Wood Bison Ratios, Survival and Fecundity Rates,
  and Abundance Estimates
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bisonpicmethod}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(bisonpictools)
library(knitr)
```

## Introduction to the Study System

Alberta Environment and Parks wishes to monitor the status of the Ronald Lake wood bison herd in northern Alberta, Canada.
A combination of annual aerial surveys and camera traps along the herd's migration corridor have been used to collect data on the status of the herd.
The aerial surveys take place once a year, usually near the end of March.
These surveys produce estimates of the total herd size, as well as the proportion of the herd that are calves.
The camera traps collect data year-round, recording photograph**(s?)** upon the detection of nearby movement.
Researchers are able to classify the individuals in the photographs captured during a camera trap "event", here defined as a **short-term series of photographs of the same group on the same day, into classes by age and sex**, as shown in Table 1. 
In the text to follow, "bull(s)" will refer to the set of $M2$, $M3$, and $MA$ individuals.

```{r, echo = FALSE, caption = "Description of the classification of Wood Bison individuals from camera trap events"}
classes <- data.frame(
  Class = c("F0","F1","FA","M0","M1","M2","M3","MA","U0","U1","UA","FU","MU","UU"),
  Description = c(
    "Female calves",
    "Female yearlings",
    "Female adults (includes 2+year-olds)",
    "Male calves",
    "Male yearlings",
    "Male 2-year-olds",
    "Male 3-year-olds",
    "Male 4+year-olds",
    "Unknown sex calves",
    "Unknown sex yearlings",
    "Unknown sex adults",
    "Unknown age females",
    "Unknown age males",
    "Individuals of unknown age and sex"
  )
)
knitr::kable(classes)
```

## Modelling approach

### Integrated Population Model

An integrated population model (IPM) was chosen to combine stage-structured population projection matrix models with multiple sources of data into a unified framework, allowing for estimates of age- and sex-class ratios, population vital rates, and abundances, while maintaining the correct propagation of uncertainty [@schaub_integrated_2022]. 
The sources of data used in this model include the classified counts from the camera trap events, and two estimates derived from aerial surveys of the herd: a "census" estimate of the entire herd size and an estimate of the proportion of calves in the entire herd.

### BAS matrix model

Population projection matrices provide a simple way to model population dynamics through time [@newman_modelling_2014].
The population projection matrix summarizes the probabilities of transitioning from each stage to every other stage, using vital rates from demographic processes such as births, age incrementation, and survival.

In this system, there are 8 stages, corresponding to the first 8 rows in table 1.
The other classes in table 1 do not represent true stages, rather the individuals' sexes and/or ages are simply unknown.
The abundances of each stage during the $t^{th}$ study year can be described by a vector, $\mathbf{n}_t$, which is composed of the numbers of individuals in each of the stages during that study year ($F0_t$, $F1_t$, etc.).

\[\mathbf{n_t} = 
\begin{bmatrix} 
    F0_{t} \\
    F1_{t} \\
    FA_{t} \\
    M0_{t} \\
    M1_{t} \\
    M2_{t} \\
    M3_{t} \\
    MA_{t} \\
\end{bmatrix}
\]

```{r, echo = FALSE}
# Insert diagram of stages!!
# need license for the software..
```

To ease model formulation, the population projection matrix can be broken up into subprocess matrices [@newman_modelling_2014].
Each subprocess has a separate matrix: in this model, the $\mathbf{B}$ matrix describes the stochastic birth process, the $\mathbf{A}$ matrix describes the deterministic age incrementation process, and the $\mathbf{S}$ matrix describes the stochastic survival process.

To get the vector of stage-wise abundances in the next year, $\mathbf{n}_{t+1}$, we multiply the subprocess matrices together in reverse chronological order, and then by the vector of stage-wise abundances in the current year, $\mathbf{n}_{t}$. 
In this model, we assume that survival occurs first, then age incrementation, and finally births, resulting in the $\mathbf{BAS}$ multiplication order:

\[\mathbf{n}_{t+1} = \mathbf{BAS}\mathbf{n}_{t}\]

With the full population abundance vectors and subprocess matrices, we arrive at the following:

\[
\begin{bmatrix}
    F0_{t+1} \\
    F1_{t+1} \\
    FA_{t+1} \\
    M0_{t+1} \\
    M1_{t+1} \\
    M2_{t+1} \\
    M3_{t+1} \\
    MA_{t+1} \\
\end{bmatrix}
= 
\begin{bmatrix}
    0 & 0 & \frac{fp_f}{2} & 0 & 0 & 0 & 0 & 0 \\
    0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
    0 & 0 & \frac{fp_f}{2} & 0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
    0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 \\
\end{bmatrix} 
\times
\begin{bmatrix}
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    0 & 1 & 1 & 0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
    0 & 0 & 0 & 0 & 0 & 0 & 1 & 1 \\
\end{bmatrix} 
\times
\begin{bmatrix}
    \phi_{F0}&0&0&0&0&0&0&0 \\
    0 & \phi_{F1} & 0 & 0 & 0 & 0 & 0 & 0 \\
    0 & 0 & \phi_{FA} & 0 & 0 & 0 & 0 & 0 \\
    0 & 0 & 0 & \phi_{M0} & 0 & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 & \phi_{M1} & 0 & 0 & 0 \\
    0 & 0 & 0 & 0 & 0 & \phi_{Bull} & 0 & 0 \\
    0 & 0 & 0 & 0 & 0 & 0 & \phi_{Bull} & 0 \\
    0 & 0 & 0 & 0 & 0 & 0 & 0 & \phi_{Bull} \\
\end{bmatrix} 
\times
\begin{bmatrix}
    F0_{t} \\
    F1_{t} \\
    FA_{t} \\
    M0_{t} \\
    M1_{t} \\
    M2_{t} \\
    M3_{t} \\
    MA_{t} \\
\end{bmatrix}
\]

where $f$ represents the fecundity rate, $p_f$ represents the proportion of fecund $FA$ individuals, and each of the $\phi$ terms represent the annual survival rate of their respective stage.
$M2$s, $M3$s, and $MA$s all share a survival rate, $\phi_{Bull}$.
Each of the $\mathbf{n}_t$ stage-wise abundance vectors represent the *expected* number of individuals in each stage in that year. 

Diffuse priors were used to provide the model with stage-wise abundances in the initial year as a starting point.
From there, the population projection matrix was used to update the stage-wise abundances in each subsequent year.
The calving season for the wood bison herd occurs between April and June, so the study year was set to begin on April 1 of each year.
As such, the above process occurs at midnight on March 31 of each study year.

Observations from camera traps occur throughout the study year, but the above stage-wise abundance and proportion vectors are the expected values on April 1.
To ensure the correct abundances and proportions were used to model this continuous data, the expected abundances and proportions were calculated for the day of the year of each camera trap event, using an exponential relationship as function of the day of the study year of the $i^{th}$ camera trap event, $d_i$:

\[\phi_{j,i} = (\phi_{j,t})^\frac{d_i}{365.25}\]

The abundances were updated for each camera trap event using a calculation similar to the projection matrix above:

\[\mathbf{n_i} = \mathbf{S_in_t}\]

where $\mathbf{n}_t$ is the stage-wise abundance vector in the study year of the $i^{th}$ camera trap event, and $\mathbf{S_i}$ is the survival matrix with the updated survival rates:

\[
\mathbf{S_i} =
\begin{bmatrix}
   \phi_{F0_i}&0&0&0&0&0&0&0 \\
   0 & \phi_{F1_i} & 0 & 0 & 0 & 0 & 0 & 0 \\
   0 & 0 & \phi_{FA_i} & 0 & 0 & 0 & 0 & 0 \\
   0 & 0 & 0 & \phi_{M0_i} & 0 & 0 & 0 & 0 \\
   0 & 0 & 0 & 0 & \phi_{M1_i} & 0 & 0 & 0 \\
   0 & 0 & 0 & 0 & 0 & \phi_{Bull_i} & 0 & 0 \\
   0 & 0 & 0 & 0 & 0 & 0 & \phi_{Bull_i} & 0 \\
   0 & 0 & 0 & 0 & 0 & 0 & 0 & \phi_{Bull_i} \\
\end{bmatrix} 
\]

By dividing the expected abundance of the $j^{th}$ stage on the date of the $i^{th}$ camera trap event, $\mathbf{n}_{j,i}$ by the total expected abundance on the day the $i^{th}$ event (i.e., the sum of the stage-wise abundances), we can derive the expected proportion of individuals in each stage on that day. The proportion of the population represented by the $j^{th}$ stage during the $i^{th}$ event, $p_{j,i}$ is:

\[p_{j,i} = \frac{\mathbf{n}_{j,i}}{\sum_{j = 1}^J \mathbf{n}_{j,i}}\]

where $J$ is the total number of stages (8).
The vector of these proportions, $\mathbf{p}_i$, will be used to model the data from camera trap events and the estimated proportion of calves from aerial surveys.
By rule, each vector of $\mathbf{p}_i$ values sums to 1.

### Camera Trap Event Data

Each camera trap event represents a random draw from the entire population, with the same expected proportions in the small mixed groups captured in the cameras as in the overall population.

Camera trap observations that occur during the calving season are excluded from the analysis to avoid bias associated with **_____.**
To ensure that the survival rates of the calves, $\phi_{F0}$ and $\phi_{M0}$, represent the true annual survival rates, a probability of initial mortality of calves was incorporated. **make active voice**
This allows ___ (reason) **source.**

$MA$ individuals are known to go off on their own or form small bachelor groups **source**.
This was accounted for in the model by assuming that only a proportion of MAs are present in the mixed groups, and that this proportion varies seasonally.
This proportion will be denoted by $r_s$ for the $s^{th}$ season.
The $MA$s are expected to spend more time with the other members of the herd in __ (season) than __ (other season) **source**.
Camera trap events that detect only bulls are excluded from the analysis, because they do not represent a random draw of individuals from the total population.

An ideal way to model the camera trap data would be to use a multinomial distribution to model the probabilities that the individuals captured during a camera trap belong to each of the stages.
However, a considerable percentage of individuals (~20%) were not sexed, and were only classified as unknown calf ($U0$), unknown yearling ($U1$), or unknown adult ($UA$).
Instead, a combination of count and ratio data was used, in order to incorporate the valuable information contained in these semi-classified individuals.
The individuals that were sexed but not aged ($FU$, $MU$; ~0.01%), or neither sexed nor aged ($UU$; ~6%), were excluded from the analysis.

#### Counts

The total group size observed during event $i$ will be referred to as $N_i$, excluding individuals with unknown age ($FU$, $MU$, and $UU$).

\[N_i = F0_i + F1_i + F2_i + M0_i + M1_i + M2_i + M3_i + MA_i + U0_i + U1_i + UA_i\]

To incorporate the unsexed individuals in each age class, we add all individuals of each age class together, getting the expected number of calves, $C_i$, the number of yearlings, $Y_i$, and the number of adults, $A_i$, during the $i^{th}$ event:

\[C_i = F0_i + M0_i + U0_i\]
\[Y_i = F1_i + M1_i + U1_i\]
\[A_i = FA_i + M2_i + M3_i + MA_i + UA_i\]

The proportion of individuals from each stage expected during the $i^{th}$ event can also be added together to get the total expected proportions of calves, $p_{C_i}$, yearlings, $p_{Y_i}$, and adults, $p_{A_i}$:

\[p_{C_i} = p_{F0_i} + p_{M0_i}\]
\[p_{Y_i} = p_{F1_i} + p_{M1_i}\]
\[p_{A_i} = p_{FA_i} + p_{M2_i} + p_{M3_i} + (p_{MA_i}r_{s_i})\]

where $r_{s_i}$ is is the expected proportion of males present, in the season ($s$) of the $i^{th}$ camera trap event.

Each of the counts ($C_i$, $Y_i$, and $A_i$) then follow binomial distributions, with size equal to the group size at event $i$, $N_i$, and the probability equal to the respective proportions described above. 

\[C_i \sim \text{Binomial}(N_i, p_{C_i})\]
\[Y_i \sim \text{Binomial}(N_i, p_{Y_i})\]
\[A_i \sim \text{Binomial}(N_i, p_{A_i})\]

#### Ratios

The ratios of certain stages were also included in the model in order to incorporate the other information included in the fully-classified (i.e., known sex and age) count data from the camera trap events.

For calves and yearlings, there are two stages, one for each sex, allowing us to model the sex ratios with a binomial distribution:

\[F0_i \sim \text{Binomial}\Biggl(F0_i + M0_i, \frac{p_{F0_i}}{p_{F0_i} + p_{M0_i}}\Biggr)\]
\[F1_i \sim \text{Binomial}\Biggl(F1_i + M1_i, \frac{p_{F1_i}}{p_{F1_i} + p_{M1_i}}\Biggr)\]

Three ratios were picked to ensure differences between all four of the adult stages were accounted for.
These were the $FA$:Bull ratio, the $M2:M3$ ratio, and the $MA:FA$ ratio:

\[FA_i \sim \text{Binomial}\Biggl(FA_i + M2_i + M3_i + MA_i, \frac{p_{FA_i}}{p_{FA_i} + p_{M2_i} + p_{M3_i} + (p_{MA_i}r_{s_i})}\Biggr)\]
\[M2_i \sim \text{Binomial}\Biggl(M2_i + M3_i, \frac{p_{M2_i}}{p_{M2_i} + p_{M3_i}}\Biggr)\]
\[MA_i \sim \text{Binomial}\Biggl(MA_i + FA_i, \frac{p_{MA_i}r_{s_i}}{p_{MA_i}r_{s_i} + p_{FA_i}}\Biggr)\]

#### Covariance Structure

Wood bison often travel in groups across the landscape, particularly when migrating between their summer __ grounds and the northern calving grounds.
As such, the counts and ratios of the events can be expected to be correlated through time and space. 

To account for this, a covariance structure was implemented into each of the binomial distributional statements above. 
At a high level, this involved incorporating a random effect for each camera trap location and week, using a multivariate Normal prior with a covariance matrix that allows the covariance between observations to decline exponentially with the squared distance (in space and time) between them.
The reader is referred to Chapter 14.5.1 of @mcelreath_statistical_2020 for more detail.

### Census Data

The census data required by this portion of the model include the census estimate and an associated coefficient of variation (CV), and the day, month, and year of the aerial survey used to inform the census estimate.

To ensure the model relates correctly to the day the census estimate was completed, a process identical to that described above for the camera trap event data was done to calculate the expected stage-wise abundances and proportions for the day of year of the $a^{th}$ aerial survey.

The census data was incorporated using a normal distribution as follows,

\[C_a \sim \text{Normal}\biggl(\sum_{j=1}^J \mathbf{n}_{j,a}, \sigma_{C_a}\biggr) \space T[0, ]\]

where $C_a$ is the census estimate in the $a^{th}$ aerial survey, $\sum_{j=1}^J \mathbf{n}_{j,a}$ is the total expected abundance on the day of the $a^{th}$ study year, and $\sigma_{C_a}$ is the standard deviation of the census estimate (calculated as the product of the census estimate $C_a$ and its associated CV).
The normal distribution is lower-truncated at 0 to ensure that the abundance estimates remain positive.

### Calf Proportion Data

The data on the proportion of calves in the herd was incorporated similarly, using the expected 

\[P_a \sim \text{Normal}(p_{F0,a} + p_{M0,a}, \sigma_{P_a}) \space T[0, 1]\]

where $P_a$ is the proportion of calves estimate in the $a^{th}$ aerial survey, $p_{F0,a} + p_{M0,a}$ is the expected proportion of calves in on the day of the $a^{th}$ aerial survey, and $\sigma_{P_a}$ is the standard deviation in the estimate of the proportion of calves (calculated as the product of the proportion of calves estimate $P_a$ and its associated CV).
The normal distribution is truncated at 0 and 1 to ensure that the estimates remain on the probability scale.

### Model Fitting

The model uses

- reference to Stan, mbr, etc.?

By default, `bisonpictools` uses "report" mode and a thinning rate of 10 to save 500 MCMC samples from each of three chains (after discarding the first halves). 
To aid with debugging, other analysis modes ("debug" and "quick") can be set (see the [user guide](insert link here) for more detail).
The user can increment the thinning rate as required to achieve convergence.
The model-fitting process is automated in the [`runbisonpic`](https://github.com/poissonconsulting/runbisonpic) Shiny app.

### Predictions

Shown during the stable point of the year: March 31.
Show predictions of abundances, rates, ratios.
Derived from posterior distributions.

### Assumptions and Limitations

Key assumptions of the integrated population model include:

- The survival rates for each stage vary by study year.
- The initial mortality rate of calves varies by study year.
- The fecundity rate is fixed.
- The proportion of MA individuals present with the mixed groups varies by season.
- There is no grouping structure beyond what is accounted for by the covariance.
- Every stage is equally detectable during a camera trap event.
- Small and large groups are equally detectable.

Key limitations of the integrated population model include:

- A long run time; with ___ camera trap events, the model takes approximately __ hours to converge.
- 

## References

