---
title: "Getting Started with bisonpictools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bisonpictools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
---

<!-- Copyright 2023 Province of Alberta -->

<!-- Licensed under the Apache License, Version 2.0 (the "License"); -->
<!-- you may not use this file except in compliance with the License. -->
<!-- You may obtain a copy of the License at -->

<!-- http://www.apache.org/licenses/LICENSE-2.0 -->

<!-- Unless required by applicable law or agreed to in writing, software -->
<!-- distributed under the License is distributed on an "AS IS" BASIS, -->
<!-- WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied. -->
<!-- See the License for the specific language governing permissions and -->
<!-- limitations under the License. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(bisonpictools)
```

`bisonpictools` is an R package to facilitate the visualization and analysis of camera trap data for wood bison herds.
The package includes functions to check the correct formatting of data, visualize the data, manipulate and analyse the data using a complex custom-built Bayesian model, and generate and plot predictions of abundances, survival and fecundity rates, and population ratios.

## Installation

To install the latest development version of `bisonpictools`, execute the following code in the RStudio console.

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("poissonconsulting/bisonpictools")
```

## bisonpic Suite

`bisonpictools` is part of the bisonpic suite of tools. 
Other packages in this suite include:

- [shinybisonpic](https://github.com/poissonconsulting/shinybisonpic)
- [runbisonpic](https://github.com/poissonconsulting/runbisonpic)

## Getting Help

To get additional information on a particular function just type `?` followed by the name of the function in the R console. For example, `?bpt_analyse()` brings up the R documentation for the `bisonpictools` analysis function.

For more information on using R the reader is referred to [R for Data Science](https://r4ds.had.co.nz/) [@wickham_r_2016].

If you discover a bug in `bisonpictools` please file an [issue](https://github.com/poissonconsulting/bisonpictools/issues) with a [reprex](https://reprex.tidyverse.org/) (reproducible example).

## Data 

The easiest way to prepare the data for use with the `bisonpictools` functions is to download the templated excel spreadsheet from the [shinybisonpic website](https://poissonconsulting.shinyapps.io/shinybisonpic/).
Click the **Download Template** button, and populate each sheet with your data.

- `event_data` contains the camera trap event data, including the date and time of each event, location of the camera trap, and the number of individuals in each class
- `location_data` contains the coordinates of the camera traps
- `census_data` contains population estimates of the population from aerial surveys, and the date of the surveys
  - specifically, this should be an estimate of the entire population size, not a minimum count
- `proportion_calf_data` contains estimates of the proportion of calves in the population from aerial surveys, and the date of the surveys


## Data Visualization

### Using the RStudio Console

Functions from `bisonpictools` can be run in the RStudio console.
The `bpt_plot_ratios()` function plots the ratios of various sex-age classes over time.
The user needs to provide the event and location data, as well as vectors of classes to form the numerator and denominator of the ratio.

#### (1) Read in data from the populated excel template

```{r, eval = FALSE}
# Change `dir` to the file path of the populated excel template on your computer
dir <- "myfilepath.xlsx"
# Install readxl to read the excel spreadsheet into R
remotes::install_github("readxl")
# Read each sheet into R
location_data <- readxl::read_xlsx(dir, sheet = "location")
event_data <- readxl::read_xlsx(dir, sheet = "event")
census_data <- readxl::read_xlsx(dir, sheet = "census")
proportion_calf_data <- readxl::read_xlsx(dir, sheet = "proportion_calf")
```

#### (2) Produce ratio plots

Data checks are completed before the analysis is run to ensure they are in the correct format. 
In the plot, each point represents an event. 
The size (area) of the point indicates the ratio while the opacity of the point indicates the number of individuals in the group.

Note that a ratio value of "Inf" (infinity) indicates that the group in a particular camera trap event has no individuals in the denominator class.
For example, if plotting the calf:cow ratio, a ratio value of "Inf" would represent an event with one or more calves and no cows.
Similarly, a ratio of "0" indicates that the ratio in a particular camera trap event has no individuals in the numerator class.
For the calf:cow ratio example, this would represent an event with one or more cows and no calves.


For example, the female calf:male calf ratio plotted over all years and locations from example data:
``` {r, fig.dim = c(8, 6)}
library(bisonpictools)
bpt_plot_ratios(
  event_data,
  location_data,
  numerator = "f0",
  denominator = "m0"
)
```

It is also possible to subset the data to include one or more camera trap locations and/or study years:

``` {r, fig.dim = c(8, 6)}
bpt_plot_ratios(
  event_data,
  location_data,
  numerator = "f0",
  denominator = "m0",
  study_years = "2019-2020",
  locations = c("LOCID1", "LOCID2")
)
```

Several age/sex classes can be combined to plot ratios of interest.
For example, the next code chunk plots the calf:(cow + calf) ratio.

The label for the ratio legend can be changed from the default ("Ratio"), to more specific label, using the `ratio_name` argument. 

```{r, fig.dim = c(8, 6)}
bpt_plot_ratios(
  event_data,
  location_data,
  numerator = c("f0", "m0", "u0"),
  denominator = c("fa"),
  ratio_name = "calf:cow"
)
```

### `shinybisonpic`

The [`shinybisonpic` Shiny app](https://poissonconsulting.shinyapps.io/shinybisonpic) can be used to explore the locations of camera traps and the the ratios of different classes of wood bison in camera trap observations in a user-friendly manner.
Refer to the [user guide](https://poissonconsulting.github.io/bisonpicsuite/articles/bisonpic-user-guide.html) for more information on how to use the `shinybisonpic` app.

## Data Analysis

### Using the RStudio Console

#### (1) Read in data from the populated excel template

```{r, eval = FALSE}
# Change `dir` to the file path of the populated excel template on your computer
dir <- "myfilepath.xlsx"
# Install readxl to read the excel spreadsheet into R
remotes::install_github("readxl")
# Read each sheet into R
location_data <- readxl::read_xlsx(dir, sheet = "location")
event_data <- readxl::read_xlsx(dir, sheet = "event")
census_data <- readxl::read_xlsx(dir, sheet = "census")
proportion_calf_data <- readxl::read_xlsx(dir, sheet = "proportion_calf")
```

#### (2) Run the analysis 

This uses the `bpt_analyse()` function.
Data checks are completed before the analysis is run to ensure they are in the correct format. 
Informative error messages will print to the console if the data do not follow the required format.
The following arguments control the number of MCMC chains, number of iterations, and the thinning rate of the model:

- `analysis_mode` controls the number of iterations and chains in the model run:
  - `"debug"` is used for printing out the errors if the model does not sample (10 iterations from 2 chains),
  - `"quick"` is used for running through a quick run of the model for demonstration purposes (10 iterations from 2 chains),
  - `"report"` is the default and is used for the full analysis (500 iterations from 3 chains).

- `nthin` controls the thinning of the MCMC samples:
  - `nthin = 1L` saves every sample; use this for `"debug"` and `"quick"` modes,
  - `nthin = 10L` saves every $10^{th}$ sample, and is the detault; use this for `"report"` mode,
  - increase `nthin` by 5 if the model does not converge in "report" mode.
  
It is recommended to:

1. Run the model in `analysis_mode = "quick"` mode with a thinning rate of `nthin = 1L` to ensure the model samples correctly.
2. If there were no errors in step 1, proceed by running the model in `analysis_mode = "report"` with a thinning rate of `nthin = 10L`, to achieve convergence. 
Ensure the table output by the model has `TRUE` in the "converged" column.
If the model did not converge, increase `nthin` by 5, and re-run the model.
3. If there were errors in step 1, proceed instead by running the model in `analysis_mode = "debug"` with a thinning rate of `nthin = 1L` to print informative error messages.
  

```{r, eval = FALSE}
# Start by running once on "quick" mode, with a thinning rate of 1
analysis <- bpt_analyse(
  event_data = event_data,
  location_data = location_data,
  census_data = census_data,
  proportion_calf_data = proportion_calf_data,
  nthin = 1L,
  analysis_mode = "quick"
)

# If no errors appear, run the model on "report" mode, with a thinning rate of 10
analysis <- bpt_analyse(
  event_data = event_data,
  location_data = location_data,
  census_data = census_data,
  proportion_calf_data = proportion_calf_data,
  nthin = 10L,
  analysis_mode = "report"
)

# If errors do appear, run the model on "debug" mode, which will provide
# informative error messages
analysis <- bpt_analyse(
  event_data = event_data,
  location_data = location_data,
  census_data = census_data,
  proportion_calf_data = proportion_calf_data,
  nthin = 1L,
  analysis_mode = "debug"
)
```

#### (3) Check that the model converged

In `"quick"` or `"debug"` mode, convergence is not expected.
In `"report"` mode, convergence is expected. 
See the `converged` column in the table printed after `bpt_analyse()` is run to assess whether or not the analysis converged.
If it says `TRUE`, the model converged. 
If it says `FALSE`, the model did not converge.

This is an example of a model run in "quick" mode that did not converge (i.e., the "converged" column reads `FALSE`).

```{r, echo = FALSE}
glance <- readRDS(
  file = system.file(package = "bisonpictools", "test-objects/glance.RDS")
)
print(glance)
```


```{r, echo = FALSE}
analysis <- readRDS(
  file = system.file(package = "bisonpictools", "test-objects/analysis.RDS")
)
```

#### (4) Coefficient table

Use the `bpt_coefficient_table()` function to print out the estimated parameters from the model run.

```{r}
coef <- bpt_coefficient_table(analysis)
print(coef)
```

#### (5) Save the analysis object

Ensure that the analysis object is saved once the model finishes running.
The `bpt_save_analysis()` function saves the analysis to the desired file path.

```{r, eval = FALSE}
# Save the analysis object.
# Change the file path to the desired directory (does not require a file extension)
bpt_save_analysis(analysis, file = "file_path/analysis")
```

To load the analysis object back into R, use the complimentary function, `bpt_load_analysis()`, using the same file path it was saved to above.

```{r, eval = FALSE}
# Load analysis object if not still in the environment, using the same file path
# it was saved to in step (4) above.
analysis <- bpt_load_analysis("file_path/analysis")
```

#### (6) Make predictions 

Predictions of the stage-wise and total abundances, survival and fecundity rates, and select ratios are derived from the posterior distributions of the estimated parameters.
Use the functions with the `bpt_predict` prefix to generate the predictions.

```{r}
# Predict total abundance
bpt_predict_abundance_total(analysis)
```

```{r}
# Predicts abundance by class:
bpt_predict_abundance_class(analysis)
```

```{r}
# Predicts survival rates by class:
bpt_predict_survival(analysis)
```

```{r}
# Predicts fecundity rate and proportion of reproductive cows:
bpt_predict_fecundity(analysis)
```

```{r}
# Predicts population ratios:
bpt_predict_ratios(analysis)
```

#### (7) Plot Predictions

The `bpt_plot_predictions()` function can be used to visualize the predictions.

```{r, fig.dim = c(7, 6)}
# Plot predicted abundances by class
bpt_plot_predictions(analysis, prediction = "abundance-class")
```

```{r, fig.dim = c(7, 6)}
# Plot total abundance
bpt_plot_predictions(analysis, prediction = "abundance-total")
```

```{r, fig.dim = c(7, 6)}
# Plot survival rates
bpt_plot_predictions(analysis, prediction = "survival")
```

```{r, fig.dim = c(7, 6)}
# Plot fecundity rates
bpt_plot_predictions(analysis, prediction = "fecundity")
```

```{r, fig.dim = c(7, 6)}
# Plot ratios
bpt_plot_predictions(analysis, prediction = "ratios")
```

### `runbisonpic`

Alternatively, launch the local data analysis app by running the following lines of code in the RStudio console.

```{r, eval = FALSE}
# Install `runbisonpic`
remotes::install_github("poissonconsulting/runbisonpic")
# Launch local app
runbisonpic::launch_runbisonpic()
```

See the [user guide](https://poissonconsulting.github.io/bisonpicsuite/articles/bisonpic-user-guide.html) for more guidance on how to use the `runbisonpic` app.

## Going beyond `bisonpictools`

The analysis object is a list that contains the model code, mcmc samples (class `mcmcr`), and other meta-data used to fit the model.
The user can interact directly with the mcmc samples using packages such as [`mcmcr`](https://github.com/poissonconsulting/mcmcr) and [`mcmcderive`](https://github.com/poissonconsulting/mcmcderive) to derive other quantities of interest.

## References

<div id="refs"></div>

```{r, results = "asis", echo = FALSE}
cat(bpt_licensing_md())
```
