---
title: "bisonpictools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bisonpictools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bisonpictools)
```


## Analysis

- `nthin` controls the thinning of the MCMC samples; 
  - `nthin = 1L` saves all samples
  - increase `nthin` if model is not converging
  
- `event_data` is the camera trap event data frame saved from the first app
- `location_data` is the location data frame saved from the first app

- `census` is the number of bison in the herd from census counts
- `census_cv` is the coefficient of variation of the census counts (standard deviation / estimate)
- `census_study_year` is the study year of the census counts (study year starts on Apr. 1)
- `census_day_of_year` is the day of the year of the census counts (first day of year is Apr. 1). 
  - E.g. if the census was on the last day of March in 2023,
  - `census_study_year` = "2022-2023"
  - `census_day_of_year` = 365L
- `proportion_calf` is the proportion of calves in the herd
- `proportion_calf_cv` is the coefficient of variation of the proportion of calves in the herd
- `proportion_calf_study_year` is the study year of the proportion of calves in the herd (study year starts on Apr. 1)
- `proportion_calf_day_of_year` is the day of the year of the proportion of calves in the herd (first day of year is Apr. 1). 
  - E.g. if the proportion of calves was estimated on the last day of March in 2022,
  - `proportion_calf_study_year` = "2021-2022"
  - `proportion_calf_day_of_year` = 365L
